---
title: "Project 2 Writeup"
author: "Max Machalek"
date: "March 6, 2018"
output: html_document
---

todo: clarify how lag was determined, window size was decided upon,

```{r}
#read the data in and store it as "cmv"
cmv<-read.csv("http://www.uwyo.edu/buerkle/compbio/statlabs/data/hcmv.data.txt")$location
```


Looking at histograms of frequencies: (takl about probabilities)

  First, 
```{r}

#plot a histogram of palindromes with bin sizes of 2500 base pairs
hist(cmv, breaks=seq(0, 230000, by=2500), xlim = c(1, 250000), ylim = c(1, 15), 
     main = "palindrome frequency by location, bin size = 2500",
     xlab = "Location",labels=T)

#paste probability of the cluster of 13 in a bin size of 2500
#based on ppois()
text(x=120000, y = 13, labels = paste("p = .00003"))

#paste probability of the cluster of 10 in a bin size of 2500
#based on ppois()
text(x=220000, y = 10, labels = paste("p = .002"))

#################################################################################

#plot a histogram of palindromes with bin sizes of 5000 base pairs
hist(cmv, breaks=seq(0, 230000, by=5000), xlim = c(1, 250000), ylim = c(1, 20), 
     main = "palindrome frequency by location, bin size = 5000",
     xlab = "Location",labels=T)
text(x=120000, y = 18, labels = paste("p = .000046"))
text(x=222000, y = 12, labels = paste("p = .01520"))

#################################################################################

#plot a histogram of palindromes with bin sizes of 1000 base pairs
hist(cmv, breaks=seq(0, 230000, by=10000), xlim = c(1, 250000), ylim = c(1, 25), 
     main = "palindrome frequency by location, bin size = 10000", 
     xlab = "Location",labels=T)


#code for getting the listed probabilites is in the next chunk.
```


Notable Clusters:
```{r}

prob13 <- 1 - ppois(12, lambda = (2500*296)/229354) 
# probability of 13 or greater occurrences of palindromes
# with histogram bins of 2500 base pairs

show(prob13)
# 3.401389e-05

prob10 <- 1 - ppois(9, lambda = (2500*296)/229354)
# probability of 10 or greater occurrences of palindromes
# with histogram bins of 2500 base pairs
show(prob10)
# 0.00187622

#################################################################################

prob18 <- 1 - ppois(18, lambda = (5000*296)/229354)
# probability of 18 or greater occurences of palindromes
# with bins of 5000 base pairs
show(prob18)
# 4.599756e-05

prob12 <- 1 - ppois(12, lambda = (5000*296)/229354)
# probability of 12 or greater occurences of palindromes
# with bins of 5000 base pairs
show(prob12)
# 0.01520251

#################################################################################

prob23 <- 1 - ppois(23, lambda = (10000*296)/229354)
# probability of 23 or greater occurences of palindromes
# with bins of 10000 base pairs
show(prob23)
# 0.003640147

prob16 <- 1 - ppois(16, lambda = (10000*296)/229354)
# probability of 16 or greater occurences of palindromes
# with bins of 10000 base pairs
show(prob23)
# 0.003640147

```


```{r}
# cmv.counts<-slide.window(hcmv, 230000, blocksize=3000, incr=400)
# plot(count ~ block.mid, data=cmv.counts, type="l")
slide.window<-function(seqdata, total.length, blocksize=500, incr=100){

  block.right<-seq(blocksize,total.length, by=incr)

  block.left<-block.right-blocksize+1

  block.mid<-(block.right+block.left)/2

  nblocks<-length(block.right)

  count<-numeric(nblocks)

  rate<-numeric(nblocks)

  

  for(i in 1:nblocks) {

    focal<-seqdata[seqdata>block.left[i] & seqdata < block.right[i]]

    count[i]<-length(focal)

    rate[i]<-count[i] / blocksize

    # alternatively

    # count[i] <- sum(seqdata>block.left[i] & seqdata < block.right[i])

  }

  data.frame(count, rate, block.left, block.mid, block.right)

}

 

plot(rate ~ block.mid, data=slide.window(cmv, 230000, blocksize=2500, incr=1000), type="l")

lines(rate ~ block.mid, data=slide.window(cmv, 230000, blocksize=5000, incr=1000), col="green")

lines(rate ~ block.mid, data=slide.window(cmv, 230000, blocksize=10000, incr=1000), col="red")

#lines(rate ~ block.mid, data=slide.window(cmv, 230000, blocksize=36000, incr=1000), col="red")
```
```{r}
#binning by 2500 base pairs
plot(0:25, dpois(0:25, lambda=2500*296/229354), type='h',
     ylab="Probability density", xlab="Number of palindromes")
abline(v = qpois(c(0.025, 0.5, 0.975), lambda=2500*296/229354), col=c("red", "blue", "red"), lwd=2)

#binning by 5000 base pairs
plot(0:25, dpois(0:25, lambda=5000*296/229354), type='h',
     ylab="Probability density", xlab="Number of palindromes")
abline(v = qpois(c(0.025, 0.5, 0.975), lambda=5000*296/229354), col=c("red", "blue", "red"), lwd=2)

#binning by 10000 base pairs
plot(0:25, dpois(0:25, lambda=10000*296/229354), type='h',
     ylab="Probability density", xlab="Number of palindromes")
abline(v = qpois(c(0.025, 0.5, 0.975), lambda=10000*296/229354), col=c("red", "blue", "red"), lwd=2)
```


Sliding window can't be bigger than bin size because then it starts grouping multiple bins in their entirety
and muddying things up. Increment of 500, 1000 include two spikes of intrigue that are not included with 1500+ increments
and we want to make sure not to exlude potential regions, but this should be noted when ranking.
```{r}
hist(cmv, breaks=seq(0, 230000, by=2500), xlim = c(1, 250000), ylim = c(1, 15), 
     main = "(fig 1) palindrome frequency by location, bin size = 2500",
     xlab = "Location", border = FALSE, col = "Grey")
lines(count ~ block.mid, data=slide.window(cmv, 230000, blocksize=2500, incr=1000), type="l", col = "black")
abline(h = 8)
```


Sliding window here shows a spike that reaches cutoff for point of interest but only if increment is under 1500.
could be a candidate but unlikely.
```{r}
hist(cmv, breaks=seq(0, 230000, by=5000), xlim = c(1, 250000), ylim = c(1, 20), 
     main = "palindrome frequency by location, bin size = 5000",
     xlab = "Location", border = FALSE, col = "Grey")
lines(count ~ block.mid, data=slide.window(cmv, 230000, blocksize=5000, incr=1000), type="l", col = "black")
abline(h = 12)
```


```{r}
hist(cmv, breaks=seq(0, 230000, by=10000), xlim = c(1, 250000), ylim = c(1, 30), 
     main = "palindrome frequency by location, bin size = 10000",
     xlab = "Location", border = FALSE, col = "Grey")
lines(count ~ block.mid, data=slide.window(cmv, 230000, blocksize=10000, incr=1000), type="l", col = "black")
abline(h = 20)
```

Can't get some of the regions to show up with 99% probability (.01) but with 95% the 135k one pops up, so 3rd place.


SPACING GRAPH
```{r}
plot(cmv[1:(296-9)], pgamma(diff(cmv, lag=9), shape=9, rate=296/229354, log=T))
```


```{r}
cmv[1:(296-7)][ which(pgamma(diff(cmv, lag=7), shape=7, 
                             rate=296/229354) < 0.05) ]

cmv[1:(296-4)][ which(pgamma(diff(cmv, lag=4), shape=4, 
                             rate=296/229354) < 0.01) ]

cmv[1:(296-5)][ which(pgamma(diff(cmv, lag=5), shape=5, 
                             rate=296/229354) < 0.01) ]

cmv[1:(296-7)][ which(pgamma(diff(cmv, lag=7), shape=7, 
                             rate=296/229354) < 0.01) ]

cmv[1:(296-9)][ which(pgamma(diff(cmv, lag=9), shape=9, 
                             rate=296/229354) < 0.01) ]
```

```{r}
```


1 - ppois(15, lambda = 2500*296/230000) # probability for 16 or more in a bin of 2500

or 

ppois(15, lambda = 2500*296/230000, lower.tail = FALSE)

CONCLUSION (CANDIDATE REGIONS, ranked)

1.Around 90000-95000 (specifically . . . .)

2. 190k-195k

3. 135k

4. 

5. 

